LCS Tool: A Computational Platform for Lagrangian Coherent Structures

### 2014-10-07 main.tex
Switch spelling to British English
J. Comp. Sci. Guide for Authors indicates article keywords are to be
in British English. For simplicity, change entire document to British
English.

\begin{abstract}
% FIXME Reviewer #5, comment #1: Abstract and introduction the objective and contribution of the paper are not clear; several points need to be clarified further.
% FIXME Reviewer #5, comment #2: For readers to quickly catch the contribution in this work, it would be better to highlight major difficulties and challenges, and original achievements to overcome them, in a clearer way in abstract and introduction.
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool. LCSs are most repelling, attracting and shearing material lines that form the centerpieces of observed tracer patterns in two-dimensional unsteady dynamical systems. LCS Tool implements the latest geodesic theory of LCSs for two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations. After a review of the underlying theory, we explain the steps and numerical methods used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples.
\end{abstract}

@@

% FIXME Reviewer #5, comment #5: Please provide information regarding practical applications of this algorithmic software tool.
Lagrangian Coherent Structures (LCSs) are evolving organizing centers of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.

\citet{haller01:_distin} proposed that ridges of the finite-time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling- and attracting-type) LCSs. A number of examples indeed support this principle\citep{peacock13:_lagran}. Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\citep{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \citet{beron-vera10:_invar_lagran}).

@@

Any orbit of \cref{eq:etafields} turns out to stretch uniformly under the flow map $F_{t_0}^t$. Specifically, any subset of an orbit of \cref{eq:etafields} increases its arclength precisely by a factor of $\lambda$. For this reason, we refer to trajectories of \cref{eq:etafields} as $\lambda$-lines. Following \citet{haller13:_coher_lagran}, we call the outermost member of a closed family of $\lambda$-lines a coherent Lagrangian vortex boundary.

@@

To avoid excessive computational times in MATLAB, we vectorize \cref{eq:velocity}, i.e., combine its right hand side evaluated over each initial point into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ODEs. We then use MATLAB's \lstinline!ode45! function to perform trajectory integration from all grid points simultaneously. This calculation typically takes five to ten minutes with default error tolerances.
% FIXME Reviewer #2, comment #3: Add memory use data based on, for example, computations for \cref{f:double gyre lambda LCS convergence}.
A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

@@

The first step is to define the position of Poincare sections in regions where closed $\lambda$-lines are expected based on a visual analysis of the orbit structure of the $\eta_\pm^\lambda$ vector field. The Poincare section is to be oriented such that the first endpoint is close to the center of the expected Lagrangian vortex, and the second endpoint is outside this vortex. There is no automated procedure implemented in LCS Tool for the positioning of Poincare sections (see however \citet{karrasch14:_autom_lagran} for a recently developed algorithm). Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A reasonable default value is 100.

The second step is to integrate the $\lambda$-lines  starting from the Poincare section to obtain the corresponding Poincare map. Integration of $\lambda$-lines is performed using the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta} over the main grid. The underlying eigenvector fields, $\xi_i(x_0)$, have generic but removable orientation discontinuities, which require monitoring and local reorientation. This process is sketched in \cref{t:variable step integration} and illustrated in \cref{f:variable step integration}. Linear interpolation is used in the interpolation of $\eta_\pm^\lambda$ within a grid element, since using higher-order interpolation would necessitate verifying that there are no orientation discontinuities beyond the four nearest grid points. We identify orientation discontinuities by checking the inner product of the $\eta_\pm^\lambda$ vectors at adjacent grid points. Rotations exceeding 90°, between two such neighboring vectors are classified as orientation discontinuities and are corrected before linear interpolation. When setting the Cauchy-Green strain tensor main grid resolution, one may find it helpful to calculate a histogram of eigenvector field rotations to ensure that all rotations are well below 90° or almost 180°.

@@

An example of a Poincare map produced from integration of the $\eta_\pm^\lambda$ field is shown in \cref{f:Poincare return map}. Most orbits will return to the Poincare section and their integration will then be stopped using the ODE event detection function of MATLAB. Some orbits may, however, deviate far from the Poincare section and do not return for any reasonable integration time. To control this behavior, we specify a maximum orbit length, \lstinline!poincareSection.orbitMaxLength!. In practice, viewing the Poincare section as the radius of a circle and setting the maximum $\lambda$-line integration length to twice the circumference gives good results.

@@

In classical, infinite-time dynamical systems, elliptic and hyperbolic invariant manifolds cannot intersect, as the trajectories in their intersection would then have to follow two different asymptotic behaviors simultaneously. In finite-time dynamical systems, hyperbolic LCSs may well intersect elliptic LCSs, with hyperbolic LCSs continuing to act as cores of mixing patterns observed in the interior of elliptic LCSs. Motivated by convention and for clarity in visualization, however, the default setting of LCS Tool removes hyperbolic LCS segments from the interior of elliptic LCSs. These removed segments can be turned back on demand, as shown in \cref{f:ocean_dataset_colortracer}.

@@

\subsection{Double gyre}
The double gyre is a model for a time-dependent two gyre system observed in geophysical flows \citep{shadden05:_defin_lagran_lyapun}. The model consists of two counter rotating sinusoidal vortices with a harmonically oscillating line in-between. Lagrangian particle motions satisfy the non-autonomous dynamical system
\begin{equation}
\begin{split}
\od{x}{t} = -\pi A \sin[\pi f(x,t)] \cos(\pi y),\\
\od{y}{t} = \pi A \cos[\pi f(x,t)] \sin(\pi y) \pd{f(x,t)}{x},\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x.
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function describing this velocity field is given in \cref{l:double gyre derivative}, specifying the right hand side of the particle ODE in a way that supports vectorized integration.

@@

\Cref{f:ocean dataset LCS} shows the positions of elliptic and hyperbolic LCSs on 22 November 2006, which is the same time instant analyzed in \citet{haller13:_coher_lagran,beron-vera13:_objec_agulh}. Here the integration time is chosen as $T = 30$ days, as opposed to 90 days in the cited references, to avoid tangling of hyperbolic LCSs. Our analysis via LCS Tool reveals five coherent eddy boundaries. The large coherent eddy at $(3,-32)$ has a non-stretching boundary, i.e., $\lambda = 1$. It corresponds to eddy \#2 in Figure~3 of \citet{beron-vera13:_objec_agulh}. Additionally, four smaller coherent eddy cores are found. They do not stay coherent over a time of 90 days. All closed orbits are stable under increased spatial resolution for the Cauchy-Green strain tensor field. Repelling LCSs (red) and attracting LCSs (blue) are superimposed over the plot of FTLE values. The hyperbolic LCSs determine the deformation of the fluid in-between coherent eddy cores.

@@

\section*{Acknowledgments}

@@

% FIXME Keywords are required to be in British English (http://www.elsevier.com/journals/journal-of-computational-science/1877-7503/guide-for-authors), so should switch entire document to British English
% Local Variables:
% ispell-dictionary: "en_US"
% End:
% LocalWords:  stretchline stretchlines incompressible gyre eigenvector eigenvectors incompressibility dataset datasets LCS LCSs Lyapunov Bickley Rossby timespan variational adaptively vortices parametrise advected strainlines minima extremal FTLE MATLAB maxima geostrophic shrinkline shrinklines Poincare altimetry

>>>>

\begin{abstract}
% FIXME Reviewer #5, comment #1: Abstract and introduction the objective and contribution of the paper are not clear; several points need to be clarified further.
% FIXME Reviewer #5, comment #2: For readers to quickly catch the contribution in this work, it would be better to highlight major difficulties and challenges, and original achievements to overcome them, in a clearer way in abstract and introduction.
We give an algorithmic introduction to Lagrangian coherent structures (LCSs) using a newly developed computational engine, LCS Tool. LCSs are most repelling, attracting and shearing material lines that form the centrepieces of observed tracer patterns in two-dimensional unsteady dynamical systems. LCS Tool implements the latest geodesic theory of LCSs for two-dimensional flows, uncovering key transport barriers in unsteady flow velocity data as explicit solutions of differential equations. After a review of the underlying theory, we explain the steps and numerical methods used by LCS Tool, and illustrate its capabilities on three unsteady fluid flow examples.
\end{abstract}

@@

% FIXME Reviewer #5, comment #5: Please provide information regarding practical applications of this algorithmic software tool.
Lagrangian Coherent Structures (LCSs) are evolving organizing centres of trajectory patterns in non-autonomous dynamical systems\citep{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\citep{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\citep{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\citep{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\citep{gawlik09:_lagran}, crowd dynamics\citep{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\citep{hadjighasem13:_detec_kam}.

\citet{haller01:_distin} proposed that ridges of the finite-time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling and attracting type) LCSs. A number of examples indeed support this principle\citep{peacock13:_lagran}. Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\citep{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \citet{beron-vera10:_invar_lagran}).

@@

Any orbit of \cref{eq:etafields} turns out to stretch uniformly under the flow map $F_{t_0}^t$. Specifically, any subset of an orbit of \cref{eq:etafields} increases its arc length precisely by a factor of $\lambda$. For this reason, we refer to trajectories of \cref{eq:etafields} as $\lambda$-lines. Following \citet{haller13:_coher_lagran}, we call the outermost member of a closed family of $\lambda$-lines a coherent Lagrangian vortex boundary.

@@

To avoid excessive computational times in MATLAB, we vectorize \cref{eq:velocity}, i.e., combine its right hand side evaluated over each initial point into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ordinary differential equations. We then use MATLAB's \lstinline!ode45! function to perform trajectory integration from all grid points simultaneously. This calculation typically takes five to ten minutes with default error tolerances.
% FIXME Reviewer #2, comment #3: Add memory use data based on, for example, computations for \cref{f:double gyre lambda LCS convergence}.
A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

@@

The first step is to define the position of Poincare sections in regions where closed $\lambda$-lines are expected based on a visual analysis of the orbit structure of the $\eta_\pm^\lambda$ vector field. The Poincare section is to be oriented such that the first endpoint is close to the centre of the expected Lagrangian vortex, and the second endpoint is outside this vortex. There is no automated procedure implemented in LCS Tool for the positioning of Poincare sections (see however \citet{karrasch14:_autom_lagran} for a recently developed algorithm). Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A reasonable default value is 100.

The second step is to integrate the $\lambda$-lines  starting from the Poincare section to obtain the corresponding Poincare map. Integration of $\lambda$-lines is performed using the $\eta_\pm^\lambda$ vector fields defined in \cref{eq:eta} over the main grid. The underlying eigenvector fields, $\xi_i(x_0)$, have generic but removable orientation discontinuities, which require monitoring and local reorientation. This process is sketched in \cref{t:variable step integration} and illustrated in \cref{f:variable step integration}. Linear interpolation is used in the interpolation of $\eta_\pm^\lambda$ within a grid element, since using higher-order interpolation would necessitate verifying that there are no orientation discontinuities beyond the four nearest grid points. We identify orientation discontinuities by checking the inner product of the $\eta_\pm^\lambda$ vectors at adjacent grid points. Rotations exceeding 90°, between two such neighbouring vectors are classified as orientation discontinuities and are corrected before linear interpolation. When setting the Cauchy-Green strain tensor main grid resolution, one may find it helpful to calculate a histogram of eigenvector field rotations to ensure that all rotations are well below 90° or almost 180°.

@@

An example of a Poincare map produced from integration of the $\eta_\pm^\lambda$ field is shown in \cref{f:Poincare return map}. Most orbits will return to the Poincare section and their integration will then be stopped using the ordinary differential equation event detection function of MATLAB. Some orbits may, however, deviate far from the Poincare section and do not return for any reasonable integration time. To control this behaviour, we specify a maximum orbit length, \lstinline!poincareSection.orbitMaxLength!. In practice, viewing the Poincare section as the radius of a circle and setting the maximum $\lambda$-line integration length to twice the circumference gives good results.

@@

In classical, infinite-time dynamical systems, elliptic and hyperbolic invariant manifolds cannot intersect, as the trajectories in their intersection would then have to follow two different asymptotic behaviours simultaneously. In finite-time dynamical systems, hyperbolic LCSs may well intersect elliptic LCSs, with hyperbolic LCSs continuing to act as cores of mixing patterns observed in the interior of elliptic LCSs. Motivated by convention and for clarity in visualization, however, the default setting of LCS Tool removes hyperbolic LCS segments from the interior of elliptic LCSs. These removed segments can be turned back on demand, as shown in \cref{f:ocean_dataset_colortracer}.

@@

\subsection{Double gyre}
The double gyre is a model for a time-dependent two gyre system observed in geophysical flows \citep{shadden05:_defin_lagran_lyapun}. The model consists of two counter rotating sinusoidal vortices with a harmonically oscillating line in-between. Lagrangian particle motions satisfy the non-autonomous dynamical system
\begin{equation}
\begin{split}
\od{x}{t} = -\pi A \sin[\pi f(x,t)] \cos(\pi y),\\
\od{y}{t} = \pi A \cos[\pi f(x,t)] \sin(\pi y) \pd{f(x,t)}{x},\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x.
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function describing this velocity field is given in \cref{l:double gyre derivative}, specifying the right hand side of the particle ordinary differential equation in a way that supports vectorized integration.

@@

\Cref{f:ocean dataset LCS} shows the positions of elliptic and hyperbolic LCSs on 22 November 2006, which is the same time instant analysed in \citet{haller13:_coher_lagran,beron-vera13:_objec_agulh}. Here the integration time is chosen as $T = 30$ days, as opposed to 90 days in the cited references, to avoid tangling of hyperbolic LCSs. Our analysis via LCS Tool reveals five coherent eddy boundaries. The large coherent eddy at $(3,-32)$ has a non-stretching boundary, i.e., $\lambda = 1$. It corresponds to eddy \#2 in Figure~3 of \citet{beron-vera13:_objec_agulh}. Additionally, four smaller coherent eddy cores are found. They do not stay coherent over a time of 90 days. All closed orbits are stable under increased spatial resolution for the Cauchy-Green strain tensor field. Repelling LCSs (red) and attracting LCSs (blue) are superimposed over the plot of FTLE values. The hyperbolic LCSs determine the deformation of the fluid in-between coherent eddy cores.

@@

\section*{Acknowledgements}

@@

% Local Variables:
% ispell-dictionary: "en_GB"
% End:
% LocalWords:  stretchline stretchlines incompressible gyre eigenvector eigenvectors incompressibility dataset datasets LCS LCSs Lyapunov Bickley Rossby timespan variational adaptively vortices parametrise advected strainlines minima extremal FTLE MATLAB maxima geostrophic shrinkline shrinklines Poincare altimetry zonal
